The effect of clumpy outflows and disc anisotropy on quasar unification scenarios

### 2015-07-16 paper_mnras.tex
some more references and text, mnras script

%%%% Various options for document class.
%%\documentclass[usenatbib, a4paper, 11pt]{aastex}
%%\documentclass[preprint, 11pt, a4paper]{aastex}
%%\documentclass[twocolum]{revtex4}
%%\documentclass{report}
\documentclass[useAMS,usenatbib]{mn2e_x}

%\usepackage{psfig,morefloats,url}
%use preprint2 for 2 columns paper.

@@

\author[Matthews et al.]{J.~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, C.~Knigge,$^1$
N.~Higginbottom,$^1$ K.~S.~Long,$^2$ S.~A.~Sim,$^3$ and S.~W.~Mangham
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK}

@@

\maketitle

\begin{abstract} 
% Broad absorption lines (BALs) in the ultraviolet 
% are seen in $\sim20\%$ of quasi-stellar objects (QSOs). 
% Blue-shifted broad absorption lines (BALs) are the most direct evidence of 
% accretion disc `winds' in such systems; mass loaded outflows
% emanating from the disc that may be driven by line forces or
% magnetic processes. 
Various unification schemes for
quasars and luminous active galactic nuclei (AGN) have proposed
that the broad emission line region is roughly cospatial
with broad absorption line (BAL) gas and much of the phenomenology of luminous AGN
can be explained by a simple geometrical picture involving an accretion
disc and associated outflow. Here, we test this paradigm by 
utilising our state-of-the-art radiative transfer code to produce synthetic spectra
from simple biconical disc wind models. In particular, we expand on our previous 
work in which a benchmark model for BAL quasars was produced. 
We find that a simple treatment of clumping (`microclumping') 
allows for a more realistic X-ray luminosity in the model by lowering the 
ionization parameter. We examine the X-ray properties of this new model
and find good agreement with existing X-ray samples of AGN and QSOs.
We find that clumping enhances the H recombination and collisionally excited resonance lines,
causing strong line emission (EW=?) to emerge at the low inclination angles, 
which represent quasars within this unification scenario.
% We also treat Hydrogen using a `macro-atom' approach in order to 
% examine the effect of recombination on Hydrogen emission lines, and
% this results in significant line emission in Lyman~$\alpha$ and the Balmer series.
However, we are unable to produce line emission with comparable equivalent widths
to existing quasar composites, due to a fundamental constraint arising
from the anisotropy of emission from a classical thin disc. We briefly 
explore the effect of relativistic beaming, gravitational redshift and 
light bending on the angular distribution of disc continuum emission. 
We find that these general relativistic effects do cause the disc to
emit more isotropically, but this is not yet sufficient to produce a
self-consistent model. We discuss a number of potential solutions. 
Overall, our work suggests that geometric unification
involving an accretion disc wind is a promising scenario, but our results 
pose a number of difficult challenges to such a model.
% Determining the true geometry of ADWs and uncovering the true disc spectral 
% (and angular) energy distribution are key next steps if we are to build up a 
% holistic picture of the quasar population.
\end{abstract}

@@

The spectral energy distributions (SEDs) of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines (e.g. \la, \civ, \nv) 
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to emission from a geometrically thin, 
optically thick accretion disc surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars (REFs). These outflows can take the form of 
highly collimated radio jets \citep{bellonijet2010}(BETTER REF), 
or mass-loaded `winds' emanating from the accretion disc. 
Approximately $20\%$ of quasars exhibit broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within this paradigm, the BALQSO fraction is associated with
the covering factor of the outflow.

@@

Despite the clear importance of ADWs in understanding AGN SEDs and accretion physics, 
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms for ADWs have been proposed, including
thermal pressure (REFs), magnetocentrifugal forces \citep{blandfordpayne} and 
radiation pressure on spectral lines \citep[`line-driving'][]{lucysolomon1970, MCGV95}.
Of these, line-driving is possibly the most attractive, due 
to the strong absorption lines seen in BALQSOs and the X-ray spectra of AGN (REFs).
The presence of line-locked features (REFs) and the so-called `ghost of \la' \citep{arav1996}
in the spectra of such objects also gives clearer evidence that line-driving is
at least partially contributing to the acceleration of the wind.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma. \cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
Hydrodynamic simulations of line-driven disc winds also found a shielding region
was required to maintain the correct ionization state \citep{PSK2000,PK04}. 
However, \cite{H14} showed that including multiple scattering means the ionizing radiation 
field could still reach the previously shielded regions in those particular models.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/wind.eps}
\caption
{
Physical properties of the outflow, shown by the coloured contours.
The solid black line marks a sphere at $1000~R_G$.
The dotted lines show the $75^\circ$ and $80^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
}
\label{fig:uvspec}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figures/uvspec.eps}
\caption
{
Synthetic spectra at four viewing angles in the clumpy model. Plot would look different probably,
and may show composites, but this would be the main synthetic spectrum plot.
}
\label{fig:uvspec}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/opt.eps}
\caption
{
Figure designed to show optical spectrum and comparison to composite.
}
\label{fig:uvspec}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/alpha_ox_both.eps}
\caption
{
X-ray properties of the H13 and clumped model (text filled 
squares), plotted against monochromatic luminosity 
at 2500\AA. Also plotted are the samples considered by
Saez et al. 2012 on a similar plot; The COMBI-7 AGN sample (ref),
the BQS sample (ref) and the Saez et al. (2012) sample of BALQSOs.
}
\label{fig:xray}
\end{figure*}

@@

Although the low inclination, `BALQSO-like' viewing angles show moderate agreement with the data,
it appears that our models tend to underpredict the emergent X-ray luminosity at 
although we are limited by poor statistics. It appears that our models tend
It is possible that this is due to our wind being overly optically thick to 
electron scattering at angles which look through the wind. It could also be that the
isotropic X-ray source assumption is incorrect and BALQSOs are {\em intrinsically} 
X-ray weak, as suggested by REFS. Nevertheless, our input X-ray spectrum
now reproduces the X-ray properties of a luminous quasar.

We can also examine the emergent X-ray spectra. 
Figure 6 shows the broadband SED from the model, compared to a number of BALQSO spectra
from Grupe \& Nousek (2015) - note that lower mass and super-Eddington sources
are excluded. Our code is not yet optimized for producing X-ray spectra, as we do not yet 
include all lines and levels of e.g. Fe or macro-atom treatments of 
higher order ions (cf. Sim et al. 2008, 2010). 
However, the bound-free opacity sources are complete, and comparisons of the shape
of the heavily absorbed spectra can be useful.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/c4_with_i.eps}
\caption
{
Carbon IV, Al III and Mg II line profiles with inclination for wind angles.
Showing how the line profiles change
and how LoBALs kick in at particularly high inclinations.
}
\label{fig:lobal}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/adjusted_withbals.eps}
\caption
{
$F_\lambda$ normalised to $F_{2000}$. {\bf Again, this is a placeholder- but I'm thinking some kind of comparison to composite including the adjusted continuum, showing that we can't get it exactly right}.
}
\label{fig:finalcomp}
\end{figure*}

@@

\bibliography{mybib.bib,stellar.bib,h14.bib,h13.bib}
\newpage
%\appendix
\end{document}

>>>>

\documentclass[useAMS,usenatbib]{mn2e_x}
%%%% Various options for document class.
%\usepackage{psfig,morefloats,url}
%use preprint2 for 2 columns paper.

@@

%\author[Matthews et al.]
\author[Matthews et al.]{
\parbox[t]{\textwidth}{
J.~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, C.~Knigge$^1$,
N.~Higginbottom$^1$, K.~S.~Long$^2$, S.~A.~Sim$^3$, and S.~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

@@

\maketitle
\begin{abstract} 
% Broad absorption lines (BALs) in the ultraviolet 
% are seen in $\sim20\%$ of quasi-stellar objects (QSOs). 
% Blue-shifted broad absorption lines (BALs) are the most direct evidence of 
% accretion disc `winds' in such systems; mass loaded outflows
% emanating from the disc that may be driven by line forces or
% magnetic processes. 
Various unification schemes for
quasars and luminous active galactic nuclei (AGN) have proposed
that the broad emission line region is roughly cospatial
with broad absorption line (BAL) gas and much of the phenomenology of luminous AGN
can be explained by a simple geometrical picture involving an accretion
disc and associated outflow. Here, we test this paradigm by 
utilising our state-of-the-art radiative transfer code to produce synthetic spectra
from simple biconical disc wind models. In particular, we expand on our previous 
work in which a benchmark model for BAL quasars was produced. 
We find that a simple treatment of clumping (`microclumping') 
allows for a more realistic X-ray luminosity in the model by lowering the 
ionization parameter. We examine the X-ray properties of this new model
and find good agreement with existing X-ray samples of AGN and QSOs.
We find that clumping enhances the H recombination and collisionally excited resonance lines,
causing strong line emission (EW=?) to emerge at the low inclination angles, 
which represent quasars within this unification scenario.
% We also treat Hydrogen using a `macro-atom' approach in order to 
% examine the effect of recombination on Hydrogen emission lines, and
% this results in significant line emission in Lyman~$\alpha$ and the Balmer series.
However, we are unable to produce line emission with comparable equivalent widths
to existing quasar composites, due to a fundamental constraint arising
from the anisotropy of emission from a classical thin disc. We briefly 
explore the effect of relativistic beaming, gravitational redshift and 
light bending on the angular distribution of disc continuum emission. 
We find that these general relativistic effects do cause the disc to
emit more isotropically, but this is not yet sufficient to produce a
self-consistent model. We discuss a number of potential solutions. 
Overall, our work suggests that geometric unification
involving an accretion disc wind is a promising scenario, but our results 
pose a number of difficult challenges to such a model.
% Determining the true geometry of ADWs and uncovering the true disc spectral 
% (and angular) energy distribution are key next steps if we are to build up a 
% holistic picture of the quasar population.
\end{abstract}

@@

The spectral energy distributions (SEDs) of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines (e.g. \la, \civ, \nv) 
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to emission from a geometrically thin, 
optically thick accretion disc surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.
In addition to the inflowing accreting material, 
outflows are ubiquitous in AGN
and quasars (REFs). These outflows can take the form of 
highly collimated radio jets \citep{bellonijet2010}(BETTER REF), 
or mass-loaded `winds' emanating from the accretion disc. 
Approximately $20\%$ of quasars exhibit broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within this paradigm, the BALQSO fraction is associated with
the covering factor of the outflow.

@@

Despite the clear importance of ADWs in understanding AGN SEDs and accretion physics, 
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms for ADWs have been proposed, including
thermal pressure (REFs), magnetocentrifugal forces \citep{blandfordpayne} and 
radiation pressure on spectral lines \citep[`line-driving'][]{lucysolomon1970, MCGV95}.
Of these, line-driving is possibly the most attractive, due 
to the strong absorption lines seen in BALQSOs and the X-ray spectra of AGN (REFs).
The presence of line-locked features \citep{bowler2014} and the so-called `ghost of \la' \citep{arav1996}
in the spectra of such objects also gives clearer evidence that line-driving is
at least partially contributing to the acceleration of the wind.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma. \cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
Hydrodynamic simulations of line-driven disc winds also found a shielding region
was required to maintain the correct ionization state \citep{PSK2000,PK04}. 
However, \cite{H14} showed that including multiple scattering means the ionizing radiation 
field could still reach the previously shielded regions in those particular models.

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/wind.eps}
\caption
{
Physical properties of the outflow, shown by the coloured contours.
The solid black line marks a sphere at $1000~R_G$.
The dotted lines show the $75^\circ$ and $80^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
}
\label{fig:uvspec}
\end{figure*} %fullpage

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/uvspec.eps}
\caption
{
Synthetic spectra at four viewing angles in the clumpy model. Plot would look different probably,
and may show composites, but this would be the main synthetic spectrum plot.
}
\label{fig:uvspec}
\end{figure*} %fullpage

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/opt.eps}
\caption
{
Figure designed to show optical spectrum and comparison to composite.
}
\label{fig:uvspec}
\end{figure*} %fullpage

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/alpha_ox_both.eps}
\caption
{
X-ray properties of the H13 and clumped model (text filled 
squares), plotted against monochromatic luminosity 
at 2500\AA. Also plotted are the samples considered by
Saez et al. 2012 on a similar plot; The COMBI-7 AGN sample (ref),
the BQS sample (ref) and the Saez et al. (2012) sample of BALQSOs.
}
\label{fig:xray}
\end{figure*} %fullpage

@@

Although the low inclination, `BALQSO-like' viewing angles show moderate agreement with the data,
it appears that our models tend to underpredict the emergent X-ray luminosity at 
although we are limited by poor statistics. 
It is possible that this is due to our wind being overly optically thick to 
electron scattering at angles which look through the wind. It could also be that the
isotropic X-ray source assumption is incorrect and BALQSOs are {\em intrinsically} 
X-ray weak, as suggested by REFS. Nevertheless, our input X-ray spectrum
now reproduces the X-ray properties of a luminous quasar.

Typically, BALQSOs show strong X-ray absorption with columns 
of $N_H\sim10^{23}~\rm{cm^{-2}}$ 
\citep{green1996,brandt2000,mathur2000,green2001,grupemathur2003}.
This is often cited as evidence that the BAL outflow is shielded from
the X-ray source, especially as sources with strong X-ray absorption tend
to exhibit deep BAL troughs and high outflow velocities (REFs).
Our results tend to imply that a clumpy BAL outflow
itself can be responsible for the strong X-ray absorption, and possibly 
explains why mini-BALs have weaker X-ray absorption than BALQSOs 
\citep[e.g.][]{brandt2000,hamann2013}.

We can also examine the emergent X-ray spectra from our model. 
Our code is not yet optimized for producing X-ray spectra, as we do not yet 
include all lines and levels of e.g. Fe or macro-atom treatments of 
higher order ions (cf. Sim et al. 2008, 2010). 
However, the bound-free opacity sources are complete, and comparisons of the shape
of the heavily absorbed spectra can be useful.
Figure 6 shows the broadband SED from the model, compared to a number of BALQSO spectra
from Grupe \& Nousek (2015) - note that lower mass and super-Eddington sources
are excluded.

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/c4_with_i.eps}
\caption
{
Carbon IV, Al III and Mg II line profiles with inclination for wind angles.
Showing how the line profiles change
and how LoBALs kick in at particularly high inclinations.
}
\label{fig:lobal}
\end{figure*} %fullpage

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/adjusted_withbals.eps}
\caption
{
$F_\lambda$ normalised to $F_{2000}$. {\bf Again, this is a placeholder- but I'm thinking some kind of comparison to composite including the adjusted continuum, showing that we can't get it exactly right}.
}
\label{fig:finalcomp}
\end{figure*} %fullpage

@@

\bibliography{mybib.bib,stellar.bib,h14.bib,h13.bib,hamann.bib}
\newpage
%\appendix
\end{document}
